---
title: "Simple Bayesian Inference of pop size from species 2 CKMRpop simulations"
output: 
  html_notebook:
    toc: true
    df_print: paged
  html_document:
    toc: true
    df_print: paged
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

OK!  I simulated 20 reps of the species_2 life history at population sizes of
500,000, 1 million, and 1.5 million individuals of ages 2 and up.  The output
from that is stored in the directory `./cluster_runs`.  We will use output from
each replicate to compute the posterior distribution of population size from
half-siblings that were born in different years.  

There is an interesting wrinkle that full avuncular relationships (aunt/uncle with
niece/nephew), as well as grandparent-grandoffspring have the same expected degree of genome sharing as half-siblings.  So, they could confound this estimation.  A full
accounting of this would include the probability of such relations, explicitly,
in the model, and use the unobserved counts of them in the likelihood.  We will
be developing such a model in full over the period of the grant.  However, currently,
in order to make a power analysis, we will just sample from the observed distribution
of the fraction of half-sibs + full-avuncular + grandparent-grandoffspring 
relationships are truly half-sibs, across different age gaps and sampling times,
and use that to induce the additional uncertainty in the population size estimates
that might be expected.

## The (simulated) data

In working through the likelihood model for this scenario in which a stable
age distribution over the last two generations is being assumed, it became clear that
all the information about ages and time of sampling of different pair members boil
down to different categories of the difference between the years that the two
individuals were born (call that $|b_2 - b_1|$) and also the ages at which the two
individuals were sampled (not necessarily ordered according to the years they
were born, so we can take those to be in ascending order.)

So, that turns out to be how we would like to summarize the data.  We will do that
for all the pairs related as either half-sibs (Si-1), full-avuncular (A-2), or grandparent-grandchild (GP-1).  So, let's get into that and do it!

```{r setup, message=FALSE}
library(tidyverse)
library(CKMRpop)
```

We need not only all the pairs with given $|b_2 - b_1|$ that are of our three focal 
relationships, but we also must have the total number of such pairs, because that
is required for the likelihood as well.  So, that will take a little bit of
extra haggling.

We count up the focal relationships first:
```{r}
# get the simulation outputs

files <- dir(
  path = "../cluster_runs",
  pattern = "cz_.*\\.rds",
  full.names = TRUE
)
names(files) <- basename(files)

output_list <- lapply(files, read_rds)

# here is a function to count them all up for a single simulation
count_classes <- function(x) {
  x$ds_pairs %>%
    mutate(
      syear1 = map_int(samp_years_list_1, function(x) x[1]),
      syear2 = map_int(samp_years_list_2, function(x) x[1]),
      sage1 = syear1 - born_year_1,
      sage2 = syear2 - born_year_2, # now make a parallel-ordered version of those
      osage1 = pmin(sage1, sage2),
      osage2 = pmax(sage1, sage2),
      birth_diff = abs(born_year_2 - born_year_1),
      relat_str = str_c(dom_relat, "-", max_hit)
    ) %>% 
    filter(
      relat_str %in% c("Si-1", "A-2", "GP-1")
    ) %>%
    count(birth_diff, osage1, osage2, relat_str) %>%
    pivot_wider(
      names_from = relat_str,
      values_from = n, 
      values_fill = 0L
    )
}  
# and here we count them up and store them in a list column.
# Note that we will filter these down to different birth diffs
# at some later time.
pair_cat_counts <- tibble(
    sim_num = 1:60,
    sim_id = names(output_list),
    cohort_size = map_chr(output_list, function(x) x$cohort_size),
    pair_categ_tibs = map(output_list, count_classes)
  )

```

That was the easy part.  Now we need to count the total number of pairs
in these different birth-diff by osage1_2 classes. We need to tally up
the total number of samples with particular properties, and then compute
the number of pairs between them, and then sum those up a bit.

```{r}
# a function to count the total number of pairwise comparisons
compute_sample_pairs <- function(x) {
  s1 <- x$ds_samples %>%
    mutate(
      syear = map_int(samp_years_list, ~.x[1]),
    ) %>%
    count(born_year, syear) %>%
    mutate(id = 1:nrow(.)) %>%
    select(id, born_year, syear, n)
  
  s2 <- s1
  names(s1) <- str_c(names(s1), "1")
  names(s2) <- str_c(names(s2), "2")
  
  s3 <- bind_cols(
    s1[rep(1:nrow(s1), each = nrow(s1)), ],
    s2[rep(1:nrow(s2), times = nrow(s2)), ]
  ) %>%
    filter(id1 <= id2) %>% # don't count these pairwise groups more than once
    mutate(
      n_pairs = case_when(
        id1 == id2 ~ n1 * (n1 - 1) / 2,
        TRUE ~ n1 * n2 * 1
      ),
      birth_diff = abs(born_year2 - born_year1),
      age1 = syear1 - born_year1,
      age2 = syear2 - born_year2,
      osage1 = pmin(age1, age2),
      osage2 = pmax(age1, age2)
    ) %>%
    group_by(birth_diff, osage1, osage2) %>%
    summarise(tot_pairs = sum(n_pairs), .groups = "drop")
  
  s3
}

# then get a tibble of them:
tot_pair_counts <- tibble(
    sim_num = 1:60,
    sim_id = names(output_list),
    cohort_size = map_chr(output_list, function(x) x$cohort_size),
    tot_pair_tibs = map(output_list, compute_sample_pairs)
  )
```


### Quick check on the total number of pairs

To be sure that I did not gooch the total number of pairs in each category,
we note that the sum of them all in any simulation should be equal to
$10,000(10,000 - 1) / 2 = 49,995,000$ pretty close to 50,000,000 if the sample
sizes are 10,000.

```{r}
tot_pair_counts %>% 
  unnest(tot_pair_tibs) %>% 
  group_by(sim_num) %>% 
  summarise(all_pairs = sum(tot_pairs))
```

Yep! Those are right on.  So, we have counted the total number of pairs in each category
up correctly.

### Join total pairs onto observed numbers 
Now, we join those together and unnest, so we have everything in a single data frame.  Note that
we can ignore any categories that didn't have any Si-1's, A-2's, or GP-1's, so we left join on that one.

While we are at it, we compute the fraction of the kin pairs that are not Si-1's.
```{r}
pcc_u <- pair_cat_counts %>%
  unnest(pair_categ_tibs)
tpc_u <- tot_pair_counts %>%
  unnest(tot_pair_tibs)

all_pairs_n <- left_join(
  pcc_u,
  tpc_u
) %>%
  replace_na(data = ., replace = list(`GP-1` = 0, `A-1` = 0)) %>%
  mutate(
    totKin = `Si-1` + `A-2` + `GP-1`,
    `fractNonSi-1` = (`A-2` + `GP-1`) / totKin
  )

# Here is what the first 20 rows of that look like:
all_pairs_n %>%
  slice(1:20)
```




### The distribution of A-2 and GP-1 relative to Si-1

We want to see how the fraction of A-2 and GP-1 varies across these different
birth_year differences and pair ages.  Let's make a plot to look at these
fractions:
```{r}
all_pairs_n %>%
  mutate(categ = str_c(birth_diff, osage1, osage2, sep = "-")) %>%
  ggplot(aes(x = totKin, y = `fractNonSi-1`, fill = factor(cohort_size))) + 
  geom_point(shape = 21, stroke = 0.2, size = 2) +
  facet_wrap(~ categ, ncol = 4)
```

We see that apart from a birth-diff of 0, the fraction of total kin with 
a genome sharing equivalent of Si-1 that are A-2 and GP-1 is pretty constant
across population sizes, as we might expect.  The big difference is that the
proportion is more variable as the total number of kin is smaller.  So, I 
think that for the purposes of propagating uncertainty, we can just compute
the average across all runs with birth-diff > 0 and use that as a
correction factor, but perhaps simulate the numbers of Si-1 amongst totKin as a binomial,
instead of just taking the expected fraction (or perhaps not do that second
part...)

So, what is that fraction?
```{r}
NSIF <- all_pairs_n %>%
  filter(birth_diff > 0) %>%
  summarise(nonSi_fract = sum(`A-2` + `GP-1`) / sum(totKin)) %>%
  pull(nonSi_fract)
NSIF
```

OK, we are looking at about 28% of the kin pairs that look like Si-1 actually being
one of the other categories.  That will play into our estimates.

## The Parts of the Kinship Probabilities

We want to compute the probability that a randomly drawn pair of samples, 1 and 2,
have an `Si-1` relationship given the demography and the population size.  We can derive
that probability by breaking it down as follows.

We will do all the calculations as if the pair are maternal half sibs, and then account
for paternal halfsibs at the end.  Each subsection below is a different part of the calculation.

### Probability of the age of the mother when the oldest pair member is born

First, we don't know the age of the mother of 1 and 2 at the time that the oldest
of the pair was born. Let's call that age $A$.  We will end up summing over all possibilities
weighted by their probability of occurrence.  Since we have a stable age distribution assumption
the probabilities of $A$ do not depend on the actual year.  Rather, they are constant over
time, and depend on the relative reproductive outputs of the different year classes.  
This is a function of the probability of:

- The stable age distribution (how many female fish of each age that are out there)
- The age-specific probability of reproducing.
- The age-specific relative reproductive potentials / fecundities.

(Note that in this case all these things have been assumed to be the same for
males and females, which will make it simpler in the end). So we have:
```{r}
# here is the relative expected reproductive output of a single individual
# of a given age versus a single individual of all other ages
RERO_1 <- species_2_life_history$`fem-prob-repro` * species_2_life_history$`fem-asrf`

# stable age distribution
SAD <- leslie_from_spip(species_2_life_history, 1000)$stable_age_distro_fem 
# there is some jiggering to do to make sure the year classes line up with
# the way we have parameterized reproductive success, etc.
SAD <- SAD[-1] # tweeze off the 0 class
SAD <- c(SAD, SAD[length(SAD)] * species_2_life_history$`fem-surv-probs`[length(species_2_life_history$`fem-surv-probs`)])
SAD <- SAD / sum(SAD)

# Now, the probability of a being a mother from a given age class is the
# relative reproductive output of each age class, which has to take account
# of the stable age distribution, like this:
RERO_AC <- RERO_1 * SAD / sum(RERO_1 * SAD) 
RERO_AC
```
So, those are the relative expected reproductive outputs of each age class from 1 to 20.

### Probability of the mother surviving $|b_2 - b_1|$ years after giving birth to the oldest member of the pair

To have a half-sibling pair, the mother at age $A$ must give birth to the first member of the
pair, and then it has to survive for $|b_2 - b_1|$ years. This depends on the survival probabilities.
Here we write a function giving those age-specific survival probabilities for $y = |b_2 - b_1|$ years,
starting from age $A$.
```{r}
# sp is a vector of survival probs, the first is the prob of surviving from 0 to 1.  
assp <- function(y, sp) {
  if(y == 0) return(rep(1, length(sp)))
  
  # fiddle the survival probs to get them to reflect 1-year-olds
  sp2 <- sp[-1]
  sp2 <- c(sp2, rep(0, 300))  # get the zero probability for going beyond the max age
  
  lapply(1:length(sp), function(x) prod(sp2[x:(x+y-1)])) %>%
    unlist()
  
}

# here are some examples:
SP <- species_2_life_history$`fem-surv-probs`
assp(0, SP)
assp(1, SP)
assp(2, SP)
assp(3, SP)
```


### Probability that the younger member of the pair has the same mother as the older member of the pair

Now, if the mother of age $A$ gave birth to the first member of the pair and then survived $y = |b_2 - b_1|$ years, then,
what has to happen is that the second member of the pair must be the offspring of that same individual.  This is the
part that depends on the total population size.  The probability of that same individual being the mother of the
second member is the expected reproductive output of a female aged $A+y$ divided by the expected reproductive output
of all females in the population at the time that the second member of the pair is born.
We have an added wrinkle in that we are defining the population size in terms
of the number of individuals of ages 2 and up.  So, we need to account for that, adding the 1-year-olds in for the calculations.
Basically, we need to be able to compute the number of females of each age group, according to the stable age
distribution, when the number of age $2+$ individuals is $N$. We are assuming an equal sex ratio here, so the 
distribution of the number of females, starting from age 1, given $N$ of age $2+$, and the stable age distribution can be
given by this quick little function:
```{r}
sad_fem_counts <- function(N, SAD) { 
  two_plus <- N * SAD[-1] / sum(SAD[-1])
  ones <- N * SAD[1] / sum(SAD[-1])
  c(ones, two_plus) / 2   # divide by two because we are just counting females
}
```
So, for example, if we thought there were 1 million fish age two and older, this
is the expected number of females of each age class, starting from 1:
```{r}
sad_fem_counts(1e6, SAD)
```

Armed with this, we can calculate the probability that the second member of the pair has the same mother,
when the total number of fish age 2 and up is 1 million.
```{r}
# Example first offspring born when mother is age A=2 and second after y=2 years
A <- 2
y <- 2
N <- 1e6

prob <- RERO_1[A + y] / sum(RERO_1 * sad_fem_counts(N, SAD))

prob
```
That is looking like 1.8 out of a million.  Which seems to be in the correct ballpark, as we have about 870 K females, total, when
you include the 1-year-olds there.

###  That's it for this calculation!

When I first started piecing this together, I thought that there must be a term
in there for the probability that the two individuals of the pair survived to
their respective ages in order to be sampled, but that turns out to NOT be necessary.
In making this calculation, we are effectively conditioning upon the fact that they
have survived.  Of course, if the ages of the members of the pair were uncertain, we would
end up having to sum over the probabilities given their possible different ages, each
weighted by the probability of those ages given some information about the fish 
that is related to age (like length, etc).  Not an issue here with age assumed
known, however.  


## Putting it all together into a single function

Here we use the above results to create a single function that gives us the
probability that a pair of fish are half-siblings.  We assume that in different
years you wouldn't get full-siblings at all, so the prob of being _either_
a maternal or a paternal half sibling is just the sum of the probs of both
of those outcomes, which means just twice the prob of being a maternal half-sib.

Note that I am not going to bother trying to write this function so that it is
vectorized over any of the input variables.  It just isn't worth it at this point...

I have included this function in CKMRpop for convenience,
since we will use it for a couple of different sample sizes for species 2,
but it is not really for general or ready for prime time.  Here is
what the function looks like:

```r
`r paste(readLines("../R/half_sib_kin_probs.R"), collapse = "\n")`
```





## Checking our probabilities

Now we should be able to compute these for each sampling category and compare expected to observed
number of kin pairs found.

```{r}
# first, find all the different categories
pair_cats <- all_pairs_n %>%
  distinct(cohort_size, birth_diff, osage1, osage2) %>%
  mutate(pop_size = case_when(
    cohort_size == "1110000" ~ 1.5e6,
    cohort_size == "740000" ~ 1.0e6,
    cohort_size == "370000" ~ 0.5e6,
  )) %>%
  select(cohort_size, pop_size, birth_diff, osage1, osage2)
```
Now, for each of those we can compute the half-sib probs:
```{r}
calced_probs <- pair_cats %>%
  mutate(
    hs_prob = pmap_dbl(
      .l = list(
        y = birth_diff,
        a1 = osage1,
        a2 = osage2,
        N = pop_size),
      .f = function(y, a1, a2, N) {
        half_sib_kin_probs(  
          y = y, 
          a1 = a1, 
          a2 = a2, 
          N = N,
          surv_probs = species_2_life_history$`fem-surv-probs`,
          prob_repro = species_2_life_history$`fem-prob-repro`,
          rel_repro = species_2_life_history$`fem-asrf`
        )
      }
    )
  )
```

Now, we can join those to the observed ones and see how they compare:
```{r}
comp <- all_pairs_n %>%
  left_join(calced_probs) %>%
  mutate(
    expected_Si1 = hs_prob * tot_pairs,
    categ_str = str_c(birth_diff, osage1, osage2, sep = "-")  
  ) 

comp %>%
  filter(birth_diff > 0) %>%
ggplot(., aes(x = expected_Si1, y = `Si-1`, fill = categ_str)) +
  geom_point(shape = 21, stroke = 0.2) +
  geom_abline(intercept = 0, slope = 1) +
  facet_wrap(~ pop_size, ncol = 1)

```
Cool.  That looks right on.

## Doing the actual estimation

Now, the moment we have been waiting for.  We are going to 
compute all the half-sib probs for the different birth diffs and the
different ages at sampling for a variety of N's from 50,000 to 6 million
in steps of 10,000:
```{r}
Npts <- seq(5e4, 6e6, by = 1e4)
Nprobs <- all_pairs_n %>%
  distinct(birth_diff, osage1, osage2) %>%
  mutate(
    hs_prob = pmap(
      .l = list(
        y = birth_diff,
        a1 = osage1,
        a2 = osage2
      ),
      .f = function(y, a1, a2) {
        half_sib_kin_probs(  
          y = y, 
          a1 = a1, 
          a2 = a2, 
          N = Npts,
          surv_probs = species_2_life_history$`fem-surv-probs`,
          prob_repro = species_2_life_history$`fem-prob-repro`,
          rel_repro = species_2_life_history$`fem-asrf`
        )
      }
    )
  ) %>%
  mutate(
    hs_tib = map(hs_prob, function(x) tibble(N = Npts, hs_prob = x))
  ) %>%
  select(-hs_prob) %>%
  unnest(cols = c(hs_tib))
  
```

Then, we will join the probabilities on there and we use the total kin corrected by
the non-Si-1 fraction and the tot_pairs to compute the contributions to the log likelihood.
Sweet!

```{r}
# first get the logl terms from all of them (even birth_diff == 0 ones)
logl_terms <- all_pairs_n %>%
  left_join(
    Nprobs, 
    by = c("birth_diff", "osage1", "osage2")
  ) %>%
  mutate(
    logl_term = (totKin * (1 - NSIF) * log(hs_prob)) +
      ((tot_pairs - totKin * (1 - NSIF)) * log(1 - hs_prob))
  )

# then, let's get the scaled likelihoods/posteriors out of those,
# with birth_diff > 0
scaled_likelihoods <- logl_terms %>%
  filter(birth_diff > 0) %>%
  group_by(sim_num, cohort_size, N) %>%
  summarise(
    logl = sum(logl_term)  # this is the sum over age and birth_diff categories of the binomial terms
  ) %>%
  mutate(
    normo_logl = logl - max(logl),
    scaled_likelihood = exp(normo_logl) / sum(exp(normo_logl))
  ) %>%
  mutate(pop_size = case_when(
    cohort_size == "1110000" ~ 1.5e6,
    cohort_size == "740000" ~ 1.0e6,
    cohort_size == "370000" ~ 0.5e6
  ))

# while we are at it, let's filter these down to the maxes
# as well, to throw a rug onto the plots
like_maxes <- scaled_likelihoods %>%
  filter(normo_logl == 0)
  

```


Now, we can plot those things, which are essentially posterior probability curves:
```{r}
g <- ggplot(
  scaled_likelihoods,
  aes(x = N, y = scaled_likelihood, colour =  factor(pop_size), group = sim_num)
) +
  geom_vline(aes(xintercept = pop_size)) + 
  geom_line(size = 0.2) +
  geom_rug(data = like_maxes, aes(y = NULL)) +
  facet_wrap(~ pop_size, ncol = 1, scales = "free_y") +
  xlim(0, 3e6)

dir.create("outputs/002", recursive = TRUE, showWarnings = FALSE)
ggsave("outputs/002/posterior-curves-10K-samples-total.pdf", plot = g, width = 8, height = 7)
g
```

### Coefficients of variation, etc

We can approach this from a Bayesian perspective first, calculating the
variance (and from that the standard deviation) of the posterior distribution, and dividing the
posterior mean by the standard deviation of the distribution.  
```{r}
coeffs_of_v <- scaled_likelihoods %>%
  group_by(sim_num, pop_size) %>%
  summarise(
    posterior_mean = sum(N * scaled_likelihood),
    posterior_variance = sum( scaled_likelihood * ((posterior_mean - N) ^2) ),
    posterior_sd = sqrt(posterior_variance),
    coeff_of_variation = posterior_sd / posterior_mean
  )
coeffs_of_v
```

So, that is showing that the CV of the posterior distribution for each case
is pretty similar. Averaged over the 20 runs we have:
```{r}
coeffs_of_v %>%
  group_by(pop_size) %>%
  summarise(mean_cv = mean(coeff_of_variation))
```
So, those seem quite small.

We can also look at the observed distribution of MLEs to estimate the mean-squared
error, and then take the square root of that.  And look at the ratio of the sqrt(MSE) to thetrue
value as another CV-like measurement. 
```{r}
like_maxes %>%
  group_by(pop_size) %>%
  summarise(
    MSE = mean( (pop_size - N) ^ 2),
    sqrt_MSE = sqrt(MSE),
    ratio = sqrt_MSE / pop_size[1]
  )
```

So, that is reasonably close to what we computed from the posterior distribution.

In the end, let's save the scaled likelihoods.
```{r}
scaled_likelihoods %>%
  ungroup() %>% 
  write_rds("outputs/002/scaled-likelihoods-10K.rds", compress = "xz")
```
