---
title: "Developing kelp-rockfish-like simulations"
output: 
  html_notebook:
    toc: true
    df_print: paged
  html_document:
    toc: true
    df_print: paged
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```


```{r setup, message=FALSE}
library(tidyverse)
library(CKMRpop)
```

Putting these together to investigate potential for doing multigenerational
CKMR with samples taken from different places, and trying to estimate
dispersal.  Using rough numbers.

## spip parameters



### Life history

Sampling of the juveniles is lethal.  Not necessarily so for the adults,
but the lethal-sampling flag on spip is an all or nothing thing, so we are
going to be simulating lethal sampling all around.  

We will assume a max age of 25.

There are a couple of studies on when they become reproductively mature. It looks like a safe bet will be: a few 4 year-olds, some 5-year-olds, more 6-year-olds,
most 7-year-olds, and all 8-year olds.   This is bootstrapped together from
a brief CDFW report [https://web.archive.org/web/20120813223726/http://www.dfg.ca.gov/marine/nearshorefinfish/kelprockfish.asp](https://web.archive.org/web/20120813223726/http://www.dfg.ca.gov/marine/nearshorefinfish/kelprockfish.asp).

So, we will set the probabilities of reproduction at different
ages to be something like:

age  |  prob of reproducing
-----|---------------------
1    |  0
2    |  0
3    |  0
4    | 0.2
5    | 0.5
6    | 0.7
7    | 0.9
8    | 1.0

with all later older categories being 1.0's as well.  This is obviously
an approximation, but it is in the ballpark.
```{r}
pars <- list()
pars$`max-age` <- 25
pars$`number-of-years` = 80
pars$`fem-prob-repro` <- c(0, 0, 0, 0.2, 0.5, 0.7, 0.9, 1.0, rep(1.0, 25 - 8))
pars$`male-prob-repro` <- c(0, 0, 0, 0.2, 0.5, 0.7, 0.9, 1.0, rep(1.0, 25 - 8))
```

There is probably considerable mortality of recruits before maturity,
but, because the sampling of recruits is lethal, we don't actually have to
model that.  We can model the number of recruits as the number of fish that
will actually be around when they start maturing, so we can let them all
survive, every year until the onset of maturity.  This decreases the computation somewhat.  We really don't have much of an idea of what the survival rates
are for ages 4 and up, but we will assume that it is about 50% and then it ramps
up to about 85% for the older fish.

```{r}
pars$`fem-surv-probs` <- c(1, 1, 1, 0.5, 0.56, 0.62, 0.68, 0.74, 0.8, rep(0.85, 25 - 9))
pars$`male-surv-probs` <- pars$`fem-surv-probs`
```

And, we will also let female fecundity and male reproductive
potential increase with age (so long as they are still alive, which
will not be true of most of many of the older categories).
Recall that these are all relative values.
```{r}
pars$`fem-asrf` <- c(0, 0, 0, seq(4,25))
pars$`male-asrp` <- pars$`fem-asrf`
```

And, finally, add some other life-history parameters:
```{r}
pars$`offsp-dsn` <- "negbin"
pars$`fem-rep-disp-par` <- 0.25
pars$`male-rep-disp-par` <- 0.25
pars$`sex-ratio` <- 0.5
pars$`mate-fidelity` <- 0.15 # some multiple paternity

```

Cool.  Those are basic life-history parameters.  Let's save them in a separate
variable:
```{r}
kelpy_life_history <- pars
```


### Populations/demes/sampling locations

We have 9 sampling locations, and I have broken the coastline up
into 9 regions roughly centered on these sampling locations. We will assume
that things are relatively well mixed within these regions (not a bad assumption given the 2 month larval dispersal) and pretend that our samples are a random sample from each region.

Dan estimated the number of adults (and the number of mature adults) in each
region given the extent of kelp forest.  That info looks like this:
```{r, message=FALSE}
psizes <- read_csv("inputs/002/kelp_areas_and_adults.csv")

psizes
```

Those will serve as our relative sizes for the simulations, and we will have to
use them to set dispersal rates that make sense.  In the meantime,
let's write a function to set the cohort sizes and the initial sizes
of each popululation/deme, given its target census size (of mature fish).
```{r}
#' @param N the desired number of mature fish
#' @param prf the fem-prob-repro of the population
#' @param prm the male-prob-repro of the population
set_cohorts_and_inits <- function(N, pars) {
  # first check the stable age distribution for N = 1000
  check_it <- leslie_from_spip(pars, 1000)
  
  # and count up the number of mature females and males
  # in the population.  We partially count individuals of
  # different ages according to the prob of reproducing
  size <- sum(check_it$stable_age_distro_fem * pars$`fem-prob-repro`) +
    sum(check_it$stable_age_distro_male * pars$`male-prob-repro`)
  
  # now, scale the cohort size that would give us N
  cz <- N / (size / 1000)
  
  # now, use that to get the stable age distro and the init cohort sizes
  stable <- leslie_from_spip(pars, cz)
  
  # recheck for fun
  stable_size <- sum(stable$stable_age_distro_fem * pars$`fem-prob-repro`) +
    sum(stable$stable_age_distro_male * pars$`male-prob-repro`)
  
  # rescale again:
  cz2 <- N / (stable_size / cz)
  
  final <- leslie_from_spip(pars, cz2)
  
  final_size <- sum(final$stable_age_distro_fem * pars$`fem-prob-repro`) +
    sum(final$stable_age_distro_male * pars$`male-prob-repro`)
  
  # return a list
  ret <- list()
  ret$`initial-males` <- floor(final$stable_age_distro_male)
  ret$`initial-females` <- floor(final$stable_age_distro_fem)
  
  # tell spip to use the cohort size
  ret$`cohort-size` <- paste("const", ceiling(cz2), collapse = " ")
  
  ret
}
```


We can use the above function like this:
```{r}
set_cohorts_and_inits(24000, pars)
```
So, we can expand our sim-pars
```{r}
kelpy_pars <- c(
  kelpy_life_history,
  set_cohorts_and_inits(24000, kelpy_life_history),
  `discard-all` = 0
)
```

### Sampling parameters

We will do lethal sampling of recruits, that we assume are all 1-year-olds
and then, for now, let's just sample 1% of them. And then let's sample
1% of them 
```{r}
juv_samp_fract <- 0.01
sampy_pars <- list()
sampy_pars$`lethal-sampling` <- 1
sampy_pars$`gtyp-ppn-fem-pre` <- c("75-77", juv_samp_fract, rep(0, 24)  )
sampy_pars$`gtyp-ppn-male-pre` <- sampy_pars$`gtyp-ppn-fem-pre`
```


Test this:
```{r}
kelpy_pars2 <- c(
  kelpy_pars,
  sampy_pars
)

bonk <- run_spip(kelpy_pars2)
```


# Functionizing this

OK, that worked, so let us wrap that into some functions that
will give us the ability to make all of our separate populations:
```{r}
# make a set of spip pars for a single populations/deme that will
# have a stable pop size of N mature individuals. And then we 
# have sampling fractions each year of recruits and adults.
# we assume that "adults" that we might sample are 5 years and up.
spip_pars_one_pop <- function(N, juv_samp_fract, adult_samp_fract, Y = 80) {
  pars <- list()
  pars$`max-age` <- 25
  pars$`number-of-years` = Y
  pars$`fem-prob-repro` <- c(0, 0, 0, 0.2, 0.5, 0.7, 0.9, 1.0, rep(1.0, 25 - 8))
  pars$`male-prob-repro` <- c(0, 0, 0, 0.2, 0.5, 0.7, 0.9, 1.0, rep(1.0, 25 - 8))
  pars$`fem-surv-probs` <- c(1, 1, 1, 0.5, 0.56, 0.62, 0.68, 0.74, 0.8, rep(0.85, 25 - 9))
  pars$`male-surv-probs` <- pars$`fem-surv-probs`
  pars$`fem-asrf` <- c(0, 0, 0, seq(4,25))
  pars$`male-asrp` <- pars$`fem-asrf`
  pars$`offsp-dsn` <- "negbin"
  pars$`fem-rep-disp-par` <- 0.25
  pars$`male-rep-disp-par` <- 0.25
  pars$`sex-ratio` <- 0.5
  pars$`mate-fidelity` <- 0.15 # some multiple paternity

  
  tmp <- set_cohorts_and_inits(N, pars)
  pars <- c(pars, tmp)
  
  pars$`discard-all` <- 0
  
  pars$`lethal-sampling` <- 1
  pars$`gtyp-ppn-fem-pre` <- c("75-77", juv_samp_fract, 0, 0, 0, rep(adult_samp_fract, 21))
  pars$`gtyp-ppn-male-pre` <- sampy_pars$`gtyp-ppn-fem-pre`
  
  pars
}


# test it
#boink <- spip_pars_one_pop(2000, 0.01, 0.02)
#bip <- run_spip(boink)
```

# Doing a big panmictic population

I have too much work to do to get the multi-pop version of CKMRpop working before
the grant goes in, so first I just want to do a simulation with all the kelp rockfish as one big panmictic population.
```{r}
# the total estimated pop size is 410,958 fish.  So, about half a million
totPopSize <- sum(psizes$mature_adults)

# the sampling is set up to be about 2295 juveniles and 945 adults each year for three years

# get the expected stable age distribution
one_pop <- spip_pars_one_pop(410958, 0.01, 0.02)

# from that, it looks like we would get our sample sizes with a juvenile
# numbers with a sampling fraction of this:
jsf <- 2295 / (2 * one_pop$`initial-males`[1])
jsf

# and we would get our adult sampling numbers with
asf <- 945 / (2 * sum(one_pop$`initial-males`[-(1:4)]))
asf

# so now, put it in there with the right sampling fractions
one_pop2 <-  spip_pars_one_pop(410958, jsf, asf)
```

Now, let's try running that:
```{r}
big_run <- run_spip(one_pop2)
```




# Working out the multipop versions...

Now, we will put that within a function that cycles over the populations
and makes a full list for all the functions, and does island model free migration.
```{r}
#' @param psizes the tibble of pop sizes
#' @param D the fraction of each deme that leaves as larvae to then randomly
#' enter other demes OR THEIR OWN.
#' @param TotPop How large should the total pop size be.  The indiv pop sizes
#' get scaled to fit into this.
island_model_pars <- function(psizes, R, TotPop, juv_samp_fract, adult_samp_fract) {
  # we have to do some weird calculating to make an island model. In a free
  # island model, everyone would leave their population, but some would go
  # back to it, according to the relative population sizes.  Spip doesn't
  # let the pool of dispersers go back to the population they came from, so
  # we have to remove that fraction from the group that went out.
  
  # here are the relative sizes of the populations
  rs <- psizes$mature_adults / sum(psizes$mature_adults)
  
  # here are the requested pop sizes of each deme
  Ns <- ceiling(TotPop * rs)
  
  # in a free island model if R of them go out from a deme, then
  # R * rs will go back, so it is like R - (R*rs) = R(1-rs) went out.
  Mout <- R * (1 - rs)
  
  # here are the rates of migration back in
  Min <- rs
  
  # then, cycle over the demes and make them all
  P <- nrow(psizes)  # the number of pops
  pream <- list(`num-pops` = P)
  
  pops <- list()
  for(i in 1:P) {
    tmp <- spip_pars_one_pop(N = Ns[i], juv_samp_fract, adult_samp_fract)
    tmp$`male-prob-mig-out` = c("1-80", 1, Mout[i])
    tmp$`fem-prob-mig-out` = c("1-80", 1, Mout[i])
    tmp$`male-prob-mig-out` = c("1-80", 1, Min)
    tmp$`fem-prob-mig-out` = c("1-80", 1, Min)
    tmp$`new-pop` = character(0)
    
    pops[[i]] <- tmp
  }
  
  flatten(pops)
}
```

Now, let's test this:
```{r}
impars <- island_model_pars(psizes, 1, 50000, 0.01, 0.02)

set.seed(5)
testy <- run_spip(impars, num_pops = 9)
```
